NumPy---Array Programming in Python

### 2020-01-28 body.tex
Update of proliferation section (still needs extending).

% pandas, xarray, ...
% numba?
% distributed arrays, delayed execution arrays
% GPU arrays, tensor computation libraries
% sparse arrays
% foreign language arrays

Recent years have seen a massive explosion in data science, machine learning,
and artificial intelligence.  NumPy and its API, underlying almost every tool
in the computational toolchain, has become ubiquitous.  Still, NumPy itself
could never hope to satisfy every single specialized need of the community; it
is limited in its ability to work with very large datasets, datasets split
across multiple systems, and computations that require specialized hardware.
Furthermore, pursuing solutions to these challenges is out of scope for NumPy
development.

The community's efforts to fill the resultant gaps therefore led to a
proliferation of arrays including distributed arrays---which are stored on
multiple computers, GPU arrays---which utilize graphics processing units for
fast computation, computation graphs---which postpone and combine calculations
before executing them efficiently, and sparse arrays---which typically contain
few non-zero values, and store only those in memory.

This includes popular tensor computation libraries such as
CuPy, JAX, and Apache MXNet.
PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.
% \footnote{\url{https://cupy.chainer.org/}}
% \footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}
% \footnote{\url{https://numpy.mxnet.io/}}
% \footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
% \footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}

@@

Primarily, there
exist two types of Python array objects: (a) arrays that mimic NumPy arrays but are
fundamentally different and (b) subclasses of NumPy arrays.

There has been a steady increase in the number of external libraries that
provide arrays and a NumPy-like API for manipulating them.
Ideally, operating on these arrays using NumPy would simply work, so that end
users could write code once, and would then benefit from switching between
NumPy arrays, GPU arrays, distributed arrays, and so forth, as appropriate.

@@

To facilitate \emph{interoperability}, the NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.
% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
However, general adoption has not yet been achieved mainly for compatibility
concerns.
% explain a bit what "compatibility concerns" means ...
Thus, we are actively addressing these concerns by revising and extending
these protocols with the aim of wide interoperability that spans the whole
scientific Python ecosystem up to and including domain libraries.

>>>>

% pandas, xarray, ... - leave out Pandas since it's not an array?
% distributed arrays, delayed execution arrays
% GPU arrays, tensor computation libraries
% sparse arrays
% foreign language arrays

Recent years have seen a massive explosion in data science, machine learning,
and artificial intelligence.  NumPy and its API, underlying almost every tool
in the computational toolchain, has become ubiquitous.  Still, NumPy itself
could never hope to satisfy every single specialized need of the community;
due to its in-memory data model, it is limited in its ability to work with very
large datasets, datasets split across multiple systems, and computations that
require graphics processing units (GPUs), tensor processing units (TPUs) or
other specialized hardware for performance reasons.  Pursuing solutions to
these challenges is out of scope for NumPy development.

The community's efforts to fill the resultant gaps therefore led to a
proliferation of array implementations. Each deep learning framework created
its own arrays; PyTorch, Tensorflow, Apache MXNet and JAX arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:, 1]} vs.
\code{x.loc[:, 'time']}---by Xarray.

Importantly, some projects chose to follow the NumPy API exactly---with Dask,
CuPy, Sparse and JAX the most prominent examples---while others have
NumPy-inspired semantics but chose to deviate in places.

% \footnote{\url{https://cupy.chainer.org}}
% \footnote{\url{https://jax.readthedocs.io/en/latest/jax.numpy.html}}
% \footnote{\url{https://mxnet.incubator.apache.org}}
% \footnote{\url{https://dask.org}}
% JAX has a citation, see https://github.com/google/jax#citing-jax
% Xarray has a citation, see http://xarray.pydata.org/en/stable/faq.html#how-should-i-cite-xarray 
% \footnote{\url{https://pytorch.org/tutorials/beginner/blitz/tensor\_tutorial.html}}
% \footnote{\url{https://www.tensorflow.org/tutorials/customization/basics}}

% proliferation of arrays including distributed arrays---which are stored on
% multiple computers, GPU arrays---which utilize graphics processing units for
% fast computation, computation graphs---which postpone and combine calculations
% before executing them efficiently, and sparse arrays---which typically contain
% few non-zero values, and store only those in memory.

%This includes popular tensor computation libraries such as
%CuPy, JAX, and Apache MXNet.
%PyTorch and Tensorflow provide tensor APIs with NumPy-inspired semantics.

@@

% Ralf: FIXME I don't think this is actually important - should work exactly the same
% for users whether a NumPy-compatible API is a subclass or a duck array
Primarily, there
exist two types of Python array objects: (a) arrays that mimic NumPy arrays but are
fundamentally different and (b) subclasses of NumPy arrays.

There has been a steady increase in the number of external libraries that
provide arrays and a NumPy-like API for manipulating them.
Ideally, operating on these arrays using NumPy functions or semantics would
simply work, so that end users could write code once, and would then benefit
from switching between NumPy arrays, GPU arrays, distributed arrays, and so
forth, as appropriate.

@@

To facilitate \emph{interoperability}, the NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
% FIXME: expand to 3 or more sentences discussing current efforts and
% issues with those, how did adoption go, what is being done next
%\url{https://numpy.org/neps/nep-0037-array-module.html}
Versions of these protocols have been successfully deployed.
% Ralf: Could you provide some proof of effectiveness, e.g., "This has made it
% possible to create distributed GPU arrays, enabling .... [find a reference to
% work by Peter Entschev]".
% Best reference: https://ep2019.europython.eu/talks/fX8dJsD-distributed-multi-gpu-computing-with-dask-cupy-and-rapids/
% Below an attempt to start on this:
Thanks to these developments users of Dask and NumPy arrays are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.
% spelled out on this page: https://docs.dask.org/en/latest/gpu.html
% The video cited embedded there could be a citation:i
% https://on-demand-gtc.gputechconf.com/gtcnew/sessionview.php?sessionName=s9198-dask+and+v100s+for+fast%2c+distributed+batch+scoring+of+computer+vision+workloads
However, general adoption has not yet been achieved mainly for compatibility
concerns.
% explain a bit what "compatibility concerns" means ...
Thus, we are actively addressing these concerns by revising and extending
these protocols with the aim of wide interoperability that spans the whole
scientific Python ecosystem up to and including domain libraries.
